---
title: 'Psych 254 W15 PS #1'
author: "Rhia Catapano"
date: "January 21, 2015"
output: html_document
---

This is problem set #1, practicing the packages tidyr and dplyr. 

Preliminaries
-------------

We begin by clearing the working directory and loading in a set of useful packages.

```{r prelims, message=FALSE, warning=FALSE}
rm(list=ls())
source("helper/useful.R")
library("knitr")
```


Part 1: Data cleaning
---------------------

Load in data
```{r data1}
dat <- read.csv("data/janiszewski_rep_exercise.csv")
dat_original <- dat
```

Check for duplicates
```{r}
dat[duplicated(dat),]
#there are no rows that are completely duplicated

duplicates <- dat[duplicated(dat$WorkerId),]
duplicates
#however, there are duplicate IDs (48,58, 62)

#so, we want the data frame with all of the duplicates removed
dat <- dat[!duplicated(dat$WorkerId),]
```

Check structure/data types
```{r}
str(dat)
#it looks like the 3 problematic columns are dog_cost, plasma_cost, and sushi_cost. These three should be numeric, but instead are factors.
```

Make incorrectly factored variables into numeric
```{r}
#Clean up the answers in dog_cost
dat[which(dat$Answer.dog_cost == "five hundred"),]$Answer.dog_cost <- 500
dat[which(dat$Answer.dog_cost == "2,000"), ]$Answer.dog_cost <- 2000
dat[which(dat$Answer.dog_cost == "2499.99"), ]$Answer.dog_cost <- round(as.numeric(as.character(dat[which(dat$Answer.dog_cost == "2499.99"), ]$Answer.dog_cost)))
dat$Answer.dog_cost <- as.numeric(as.character(dat$Answer.dog_cost))

#Clean up the answers in plasma_cost
dat[which(dat$Answer.plasma_cost == "4999.99"), ]$Answer.plasma_cost <- round(as.numeric(as.character(dat[which(dat$Answer.plasma_cost == "4999.99"), ]$Answer.plasma_cost)))
dat[which(dat$Answer.plasma_cost == "0.45"), ]$Answer.plasma_cost <- 4500
#in my own data sets I'd probably exclude this person instead of assuming what they meant, but based on the cleaned version, it seems like we're assuming that they meant 4500.
dat[which(dat$Answer.plasma_cost == "4,000"), ]$Answer.plasma_cost <- 4000
dat$Answer.plasma_cost <- as.numeric(as.character(dat$Answer.plasma_cost))

#Clean up answers in sushi_cost
dat[which(dat$Answer.sushi_cost == "ehight"), ]$Answer.sushi_cost <- 8
dat[which(dat$Answer.sushi_cost == "7,75"), ]$Answer.sushi_cost <- 7.75
dat$Answer.sushi_cost <- as.numeric(as.character(dat$Answer.sushi_cost))

write.csv(dat, "janiszewski_rep_cleaned_Rhia.csv")
```



Part 2: Making these data tidy
------------------------------

Load in the data
```{r data2}
dat2 <- read.csv("data/janiszewski_rep_cleaned.csv")
```

Remove columns we don't need and rename variables
```{r select}
d.tidy <- dat2 %>% 
  select(WorkerId, Input.condition, 
         starts_with("Answer")) %>%
  rename(workerid = WorkerId,
         condition = Input.condition,          
         dog_cost = Answer.dog_cost,
         plasma_cost = Answer.plasma_cost, 
         sushi_cost = Answer.sushi_cost)
```

Gather the data into a tidy data frame
```{r}
d.gathered <- d.tidy %>%
  gather(type, cost, dog_cost, plasma_cost, sushi_cost)
```

Spread into a wide format
```{r spread}
d.wide <- d.gathered %>%
  spread(type, cost)
```


Part 3: Manipulating the data using dplyr
-----------------------------------------

Histogram representing the number of people in each price range for each item.

```{r, warning=FALSE, message=FALSE}
ggplot(d.gathered, aes(x=cost, fill=type)) +
    geom_histogram(position="dodge")

#Histogram
ggplot(data=d.gathered, aes(x=cost, fill=condition)) + facet_wrap(~type, scale="free") + geom_histogram(position="dodge", bins=8)

#Density Distribution
ggplot(data=d.gathered, aes(x=cost, fill=condition)) + facet_wrap(~type, scale="free") + geom_density(alpha=0.5)
```

Based on our histogram, we can see that all participants gave a low WTP for sushi, a mid-range WTP for dog, and a high WTP for a plasma TV.

Remove duplicates using distinct from dplyr.

```{r}
d.raw <- read.csv("data/janiszewski_rep_exercise.csv")
d.unique.subs <- distinct(d.raw, WorkerId)
```

Start by using *summarise* to compute the grand mean bet.
```{r}
summarise(d.gathered, "mean cost" = mean(cost, na.rm=TRUE))
```

Summary by item and condition
```{r}
summary <- d.gathered %>% 
  group_by(type, condition) %>% 
  summarise("mean cost" = mean(cost, na.rm=TRUE))
   
kable(summary)         
```

We need the anchors, so create a new tidy version of the data set including the anchors. 

```{r}
d.tidy <- dat2 %>% 
  select(WorkerId, Input.condition, 
         starts_with("Answer"), 
         starts_with("Input")) %>%
  rename(workerid = WorkerId,
         condition = Input.condition,          
         plasma_anchor = Input.price1,
         dog_anchor = Input.price2,
         sushi_anchor = Input.price3,
         dog_cost = Answer.dog_cost,
         plasma_cost = Answer.plasma_cost, 
         sushi_cost = Answer.sushi_cost) %>%
  gather(name, cost, 
         dog_anchor, plasma_anchor, sushi_anchor, 
         dog_cost, plasma_cost, sushi_cost) %>%
  separate(name, c("item", "type"), "_") %>%
  spread(type, cost) 
```

Compute absolute percentage change in price, and 
* By computing z-scores over items.

To do the first, use the *mutate* verb to add a percent change column, then comute the same summary as before. 

```{r}
pcts <- d.tidy %>% 
  mutate(pct_change = (anchor-cost)/anchor*100) %>% 
  group_by(item, condition) %>% 
  summarise("percentchange" = mean(pct_change, na.rm=TRUE))
   
kable(pcts)         
```

Computing z-scores over items

```{r}
z.scores <- d.tidy %>% 
 mutate(change = anchor - cost) %>%
 group_by(item) %>%
 mutate(scaled = scale(change)[,1]) %>%
 ungroup()  %>%   
 group_by(item, condition, add=FALSE) %>%
 summarize("groupmeans"=mean(scaled,na.rm=TRUE) )
```


Plot the differences and see if anything happened.  
 
First the percent change:

```{r}
#qplot was removed in the most recent version of ggplot
#qplot(item, percentchange, fill=condition, 
 #     position="dodge",
  #    stat="identity", geom="bar", 
   #   data=pcts)  

ggplot(data=pcts, aes(x=item, y=percentchange, fill=condition)) + geom_bar(stat="identity", position="dodge")
```

and the z-scores:

```{r}
#qplot(item, groupmeans, fill=condition, 
 #     position="dodge",
  #    stat="identity", geom="bar", 
   #   data=z.scores)  

ggplot(data=z.scores, aes(x=item, y=groupmeans, fill=condition)) + geom_bar(stat="identity", position="dodge")
```

This replication didn't seem to work out straightforwardly. 