---
title: 'Psych 254 W15 PS #1'
author: "Rhia Catapano"
date: "January 21, 2015"
output: html_document
---

This is problem set #1, in which we hope you will practice the packages tidyr and dplyr. There are some great cheat sheets from [RStudio](http://www.rstudio.com/resources/cheatsheets/).

The data set
------------

This data set comes from a replication of [Janiszewski and Uy (2008)](http://dx.doi.org/10.1111/j.1467-9280.2008.02057.x), who investigated whether the precision of the anchor for a price influences the amount of adjustment.

In the data frame, the `Input.condition` variable represents the experimental condition (under the rounded anchor, the rounded anchor, over the rounded anchor). `Input.price1`, `Input.price2`, and `Input.price3` are the anchors for the `Answer.dog_cost`, `Answer.plasma_cost`, and `Answer.sushi_cost` items.

Preliminaries
-------------

We begin by clearing the working directory and loading in a set of useful packages.

```{r prelims}
rm(list=ls())
source("helper/useful.R")
```


Part 1: Data cleaning
---------------------

Load in data
```{r data1}
dat <- read.csv("data/janiszewski_rep_exercise.csv")
dat_original <- dat
```

Check for duplicates
```{r}
dat[duplicated(dat),]
#there are no rows that are completely duplicated

duplicates <- dat[duplicated(dat$WorkerId),]
duplicates
#however, there are duplicate IDs (48,58, 62)

dat <- dat[!duplicated(dat$WorkerId),]
#so, we want the data frame with all of the duplicates removed
```

Check structure/data types
```{r}
str(dat)
#it looks like the 3 problematic columns are dog_cost, plasma_cost, and sushi_cost. These three should be numeric, but instead are factors.
```

Make incorrectly factored variables into numeric
```{r}
#Clean up the answers in dog_cost
dat[which(dat$Answer.dog_cost == "five hundred"),]$Answer.dog_cost <- 500
dat[which(dat$Answer.dog_cost == "2,000"), ]$Answer.dog_cost <- 2000
dat[which(dat$Answer.dog_cost == "2499.99"), ]$Answer.dog_cost <- round(as.numeric(as.character(dat[which(dat$Answer.dog_cost == "2499.99"), ]$Answer.dog_cost)))
dat$Answer.dog_cost <- as.numeric(as.character(dat$Answer.dog_cost))

#Clean up the answers in plasma_cost
dat[which(dat$Answer.plasma_cost == "4999.99"), ]$Answer.plasma_cost <- round(as.numeric(as.character(dat[which(dat$Answer.plasma_cost == "4999.99"), ]$Answer.plasma_cost)))
dat[which(dat$Answer.plasma_cost == "0.45"), ]$Answer.plasma_cost <- 4500
#in my own data sets I'd probably exclude this person instead of assuming what they meant, but based on the cleaned version, it seems like we're assuming that they meant 4500.
dat[which(dat$Answer.plasma_cost == "4,000"), ]$Answer.plasma_cost <- 4000
dat$Answer.plasma_cost <- as.numeric(as.character(dat$Answer.plasma_cost))

#Clean up answers in sushi_cost
dat[which(dat$Answer.sushi_cost == "ehight"), ]$Answer.sushi_cost <- 8
dat[which(dat$Answer.sushi_cost == "7,75"), ]$Answer.sushi_cost <- 7.75
dat$Answer.sushi_cost <- as.numeric(as.character(dat$Answer.sushi_cost))

write.csv(dat, "janiszewski_rep_cleaned_Rhia.csv")
```



Part 2: Making these data tidy
------------------------------

Now let's start with the cleaned data, so that we are all beginning from the same place.

```{r data2}
d2 <- read.csv("data/janiszewski_rep_cleaned.csv")
```

This data frame is in *wide* format - that means that each row is a participant and there are multiple observations per participant. This data is not *tidy*.

To make this data tidy, we'll do some cleanup. First, remove the columns you don't need, using the verb `select`.

HINT: `?select` and the examples of helper functions will help you be efficient.

```{r select}
d.tidy <- select(...)
```


Try renaming some variables using `rename`. A good naming scheme is:

* consistent with case
* consistent with "." or "_" ( "_" is usually preferred)
* concise as will be comprehensible to others

Try using the `%>%` operator as well. So you will be "piping" `d %>% rename(...)`.

```{r rename}
d.tidy <- ...
```


OK, now for the tricky part. Use the verb *gather* to turn this into a *tidy* data frame.

HINT: look for online examples!

```{r gather}
d.tidy <- ...
```


Bonus problem: *spread* these data back into a wide format data frame.

```{r spread}
d.wide <- ...
```


Part 3: Manipulating the data using dplyr
-----------------------------------------

NOTE: If you generally use `plyr` package, note that they do not play nicely together so things like the rename function won't work unless you load `dplyr` after `plyr`.

As we said in class, a good thing to do is always to check histograms of the response variable. Do that now, using either regular base graphics or ggplot. What can you conclude? 

```{r}
...
```


Try also using the dplyr `distinct` function to remove the duplicate participants from the raw csv file that you discovered in part 1.

```{r}
d.raw <- read.csv("../data/janiszewski_rep_exercise.csv")
d.unique.subs <- ...
```


OK, now we turn to the actual data anlysis. We'll be using dplyr verbs to *filter*, *group*,*mutate*, and *summarise* the data.

Start by using *summarise* to compute the grand mean bet. (Note that this is the same as taking the grant mean - the value will come later. Right now we're just learning the syntax of that verb.)

```{r}
...
```


This is a great time to get comfortable with the %>% operator. In brief, %>% allows you to pipe data from one function to another. So if you would have written:

`d <- function(d, other_stuff)`

you can now write:

`d <- d %>% function(other_stufF)`

That doesn't seem like much, but it's cool when you can replace:

`d <- function1(d, other_stuff)
d <- function2(d, lots_of_other_stuff, more_stuff)
d <- function3(d, yet_more_stuff)`

with

`d <- d %>% 
  function1(other_stuff) %>%
  function2(lots_of_other_stuff, more_stuff) %>%
  function3(yet_more_stuff)`

In other words, you get to make a clean list of the things you want to do and chain them together without a lot of intermediate assignments. 

Let's use that capacity to combine summarise with *group_by*, which allows us to break up our summary into groups. Try grouping by item and condition and taking means using *summarise*, chaining these two verbs with %>%.

```{r}
...
```



OK, it's looking like there are maybe some differences between conditions, but how are we going to plot these? They are fundamentally different magnitudes from one another. 

Really we need the size of the deviation from the anchor, which means we need the anchor value. Let's go back to the data and add that in.

Take a look at this complex expression. You don't have to modify it, but see what is being done here with gather, separate and spread. Run each part (e.g. the first verb, the first two verbs, etc.) and after doing each, look at head(d.tidy) to see what they do.  

```{r}
d.tidy <- d %>% 
  select(WorkerId, Input.condition, 
         starts_with("Answer"), 
         starts_with("Input")) %>%
  rename(workerid = WorkerId,
         condition = Input.condition,          
         plasma_anchor = Input.price1,
         dog_anchor = Input.price2,
         sushi_anchor = Input.price3,
         dog_cost = Answer.dog_cost,
         plasma_cost = Answer.plasma_cost, 
         sushi_cost = Answer.sushi_cost) %>%
  gather(name, cost, 
         dog_anchor, plasma_anchor, sushi_anchor, 
         dog_cost, plasma_cost, sushi_cost) %>%
  separate(name, c("item", "type"), "_") %>%
  spread(type, cost) 
```

Now we can do the same thing as before but look at the relative difference between anchor and estimate. Let's do this two ways: 

* By computing absolute percentage change in price, and 
* By computing z-scores over items.

To do the first, use the *mutate* verb to add a percent change column, then comute the same summary as before. 

```{r}
pcts <- d.tidy %>% 
  mutate(pct_change = ... ) %>% 
  ...
```


To do the second, you will need to *group* once by item, then to *ungroup* and  do the same thing as before. NOTE: you can use group_by(..., add=FALSE) to set new grouping levels, also. 

HINT: `scale(x)` returns a complicated data structure that doesn't play nicely with dplyr. try `scale(x)[,1]` to get what you need. 

```{r}
pcts <- d.tidy %>% 
  mutate(pct_change = ... ) %>% 
  ...
```


OK, now here comes the end: we're going to plot the differences and see if anything happened. First the percent change:

```{r}
qplot(item, pct, fill=condition, 
      position="dodge",
      stat="identity", geom="bar", 
      data=pcts)  
```

and the z-scores:

```{r}
qplot(item, z, fill=condition, 
      position="dodge",
      stat="identity", geom="bar", 
      data=z.scores)  
```

Oh well. This replication didn't seem to work out straightforwardly. 

END 
===